---
title: "The Birds We Notice: Avian Biodiversity and Observer Behaviour on the University of Auckland City Campus"
author: "Mengyuan Zheng"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    fig_caption: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.align  = "center",
  fig.width  = 8,
  fig.height = 5,
  dpi        = 150
)
```

---

> **GitHub Repository:** [https://github.com/mengyuanzheng9-star/uoa-campus-biota](https://github.com/mengyuanzheng9-star/uoa-campus-biota)

---

# Introduction

Walk across the University of Auckland's city campus on any given morning and you are likely to hear a tūī before you see one. The campus sits embedded within Tāmaki Makaurau — one of New Zealand's most urbanised environments — yet supports a surprisingly diverse community of birds, from native honeyeaters to introduced sparrows. Understanding which species are detected, when, and where has practical value for campus biodiversity management, and raises deeper questions about how human observers shape the biodiversity record itself.

Citizen science platforms such as iNaturalist have transformed how biodiversity data are gathered. The [Biota of University of Auckland City Campus](https://www.inaturalist.org/projects/biota-of-university-of-auckland-city-campus) project invites observers to record any organism encountered on campus, producing a freely available, opportunistically-sampled dataset spanning multiple years. For birds — arguably the most noticed taxon — this record allows us to ask: *What drives the number of bird species detected on any given day?* Is it the season, the effort of observers, the day of the week, or some combination of these factors?

This analysis pursues three questions:

1. How has bird species richness on campus changed over time, and are there seasonal patterns?
2. Which species are most frequently recorded, and do native and introduced species show different temporal patterns?
3. Can we model daily species richness as a function of observer effort, season, and time-of-week using a generalised linear model (GLM)?

The findings speak to both campus ecology and the interpretation of opportunistic biodiversity records more broadly.

---

# Methods

## Data Acquisition

Observation data were exported from the iNaturalist project *Biota of University of Auckland City Campus* on **27 April 2026** using the iNaturalist export tool, filtering to:

- **Taxon:** Aves (birds)  
- **Quality grade:** Research grade  
- **Date range:** All available records up to export date  

The exported CSV (`observations-719197.csv`) was saved in the project repository. No location data were obscured in the downloaded export.

## Data Wrangling

```{r libraries}
# ── Core data manipulation ──────────────────────────────────────────────────
library(tidyverse)
library(lubridate)
library(janitor)

# ── Modelling ───────────────────────────────────────────────────────────────
library(MASS)
# MASS masks dplyr::select — restore immediately
select <- dplyr::select
library(broom)
library(DHARMa)

# ── Visualisation ───────────────────────────────────────────────────────────
library(patchwork)
library(ggridges)
library(viridis)
library(knitr)
library(kableExtra)
```

```{r native-lookup}
# ── Manual native/introduced status lookup ───────────────────────────────────
# Based on NZ Birds Online and iNaturalist taxonomy.
# TRUE = native to NZ; FALSE = introduced/naturalised.
# Includes parent species AND subspecies names that appear in this dataset.
native_status <- tribble(
  ~scientific_name,                                ~native,
  # ── Native species (parent + subspecies) ────────────────────────────────
  "Prosthemadera novaeseelandiae",                 TRUE,
  "Prosthemadera novaeseelandiae novaeseelandiae", TRUE,
  "Hemiphaga novaeseelandiae",                     TRUE,
  "Hemiphaga novaeseelandiae novaeseelandiae",     TRUE,
  "Rhipidura fuliginosa",                          TRUE,
  "Rhipidura fuliginosa fuliginosa",               TRUE,
  "Rhipidura fuliginosa placabilis",               TRUE,   # North Island Fantail
  "Rhipidura fuliginosa melanura",                 TRUE,   # South Island Fantail
  "Todiramphus sanctus",                           TRUE,
  "Todiramphus sanctus vagans",                    TRUE,
  "Halcyon sancta",                                TRUE,
  "Zosterops lateralis",                           TRUE,   # Silvereye (native coloniser)
  "Zosterops lateralis lateralis",                 TRUE,
  "Anthornis melanura",                            TRUE,
  "Anthornis melanura melanura",                   TRUE,
  "Petroica macrocephala",                         TRUE,
  "Gerygone igata",                                TRUE,
  "Mohoua ochrocephala",                           TRUE,
  "Anas platyrhynchos superciliosa",               TRUE,   # Grey Duck
  "Ardea novaehollandiae",                         TRUE,   # White-faced Heron
  "Porphyrio melanotus",                           TRUE,   # Pūkeko
  "Porphyrio melanotus melanotus",                 TRUE,
  "Haematopus ostralegus",                         TRUE,
  "Anthus novaeseelandiae",                        TRUE,   # NZ Pipit
  "Anthus novaeseelandiae novaeseelandiae",         TRUE,
  # ── Introduced species (parent + subspecies) ────────────────────────────
  "Passer domesticus",                             FALSE,
  "Passer domesticus domesticus",                  FALSE,
  "Columba livia",                                 FALSE,
  "Columba livia domestica",                       FALSE,
  "Turdus merula",                                 FALSE,
  "Turdus merula merula",                          FALSE,
  "Platycercus eximius",                           FALSE,
  "Platycercus eximius eximius",                   FALSE,
  "Gymnorhina tibicen",                            FALSE,
  "Gymnorhina tibicen hypoleuca",                  FALSE,
  "Gymnorhina tibicen tibicen",                    FALSE,
  "Sturnus vulgaris",                              FALSE,
  "Carduelis carduelis",                           FALSE,
  "Carduelis carduelis britannica",                FALSE,
  "Carduelis chloris",                             FALSE,
  "Fringilla coelebs",                             FALSE,
  "Fringilla coelebs gengleri",                    FALSE,
  "Anas platyrhynchos",                            FALSE,  # Mallard
  "Anas platyrhynchos platyrhynchos",              FALSE,
  "Streptopelia chinensis",                        FALSE,
  "Streptopelia chinensis tigrina",                FALSE,
  "Streptopelia roseogrisea",                      FALSE,
  "Acridotheres tristis",                          FALSE,
  "Acridotheres tristis tristis",                  FALSE,
  "Corvus frugilegus",                             FALSE,
  "Turdus philomelos",                             FALSE,
  "Turdus philomelos clarkei",                     FALSE,
  "Hirundo neoxena",                               TRUE,   # Welcome Swallow (native coloniser)
  "Hirundo neoxena neoxena",                       TRUE,
  "Hirundo tahitica",                              FALSE,
  "Ardea alba",                                    FALSE   # Great Egret (vagrant)
)
```

```{r load-data}
# ── Load raw data ─────────────────────────────────────────────────────────────
raw_path <- "observations-719197.csv"   # file lives alongside this Rmd

if (file.exists(raw_path)) {
  raw <- read_csv(raw_path, show_col_types = FALSE)
  cat("✓ Real data loaded:", nrow(raw), "rows,", ncol(raw), "columns\n")
} else {
  stop("Data file not found: ", raw_path,
       "\nPlease place observations-719197.csv in the same folder as this Rmd.")
}
```

```{r clean-data}
# ── Standardise column names ──────────────────────────────────────────────────
raw_clean <- raw |> clean_names()

# No taxon_class_name column — export was already filtered to Aves
# Confirm by checking iconic_taxon_name if present
if ("iconic_taxon_name" %in% names(raw_clean)) {
  non_bird <- raw_clean |> filter(iconic_taxon_name != "Aves")
  if (nrow(non_bird) > 0) {
    cat("Warning:", nrow(non_bird), "non-Aves rows removed.\n")
    raw_clean <- raw_clean |> filter(iconic_taxon_name == "Aves")
  }
}

# ── Resolve column names robustly ─────────────────────────────────────────────
date_col     <- intersect(c("observed_on", "observed_on_string"), names(raw_clean))[1]
observer_col <- intersect(c("user_login", "user_id"), names(raw_clean))[1]

# ── Build birds tibble ────────────────────────────────────────────────────────
birds <- raw_clean |>
  rename(observed_on_col = all_of(date_col),
         user_login      = all_of(observer_col)) |>
  mutate(
    date        = as.Date(substr(as.character(observed_on_col), 1, 10)),
    year        = year(date),
    month       = month(date, label = TRUE, abbr = TRUE),
    month_num   = month(date),
    day_of_week = wday(date, label = TRUE, abbr = TRUE, week_start = 1),
    season      = case_when(
      month_num %in% c(12, 1, 2)  ~ "Summer",
      month_num %in% c(3, 4, 5)   ~ "Autumn",
      month_num %in% c(6, 7, 8)   ~ "Winter",
      month_num %in% c(9, 10, 11) ~ "Spring"
    ),
    season  = factor(season, levels = c("Spring", "Summer", "Autumn", "Winter")),
    weekend = if_else(wday(date, week_start = 1) >= 6, "Weekend", "Weekday")
  ) |>
  # Join native status lookup
  left_join(native_status, by = "scientific_name") |>
  filter(!is.na(date), year >= 2015) |>
  distinct()

cat("Total bird observations:", nrow(birds), "\n")
cat("Date range:", as.character(min(birds$date)), "to", as.character(max(birds$date)), "\n")
cat("Unique species:", n_distinct(birds$scientific_name), "\n")
cat("Unique observers:", n_distinct(birds$user_login), "\n")
cat("Native status resolved for",
    sum(!is.na(birds$native)), "/", nrow(birds), "observations\n")
```

## Derived Variables

For modelling we aggregate to the **daily** level, computing for each calendar day with at least one record:

| Variable | Description |
|---|---|
| `species_richness` | Number of distinct species observed |
| `n_observers` | Number of unique observers (effort proxy) |
| `log_effort` | log(n_observers + 1) |
| `season` | Southern Hemisphere astronomical season |
| `weekend` | Weekend vs Weekday |
| `year_c` | Calendar year centred at first year |

```{r daily-summary}
daily <- birds |>
  group_by(date, year, month, month_num, season, weekend, day_of_week) |>
  summarise(
    species_richness = n_distinct(scientific_name),
    n_obs            = n(),
    n_observers      = n_distinct(user_login),
    .groups = "drop"
  ) |>
  mutate(
    year_c     = year - min(year),
    log_effort = log1p(n_observers)
  )

cat("Days with at least one bird observation:", nrow(daily), "\n")
cat("\nSpecies richness per day:\n")
print(summary(daily$species_richness))
```

---

# Results

## Overview of the Dataset

```{r overview-table}
overview <- tibble(
  Metric = c("Total observations", "Research-grade records",
             "Unique bird species", "Unique observers",
             "First observation", "Last observation",
             "Days with ≥1 record"),
  Value  = c(
    format(nrow(birds), big.mark = ","),
    format(sum(birds$quality_grade == "research", na.rm = TRUE), big.mark = ","),
    n_distinct(birds$scientific_name),
    n_distinct(birds$user_login),
    as.character(min(birds$date)),
    as.character(max(birds$date)),
    nrow(daily)
  )
)

kbl(overview, caption = "Table 1. Dataset overview — Biota of UoA City Campus (Aves)") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

## Most Frequently Recorded Species

```{r top-species, fig.cap="Figure 1. Most frequently recorded bird species on the UoA city campus. Native NZ species are shown in teal; introduced species in salmon. Species with unresolved status shown in grey."}
top20 <- birds |>
  count(scientific_name, common_name, native, sort = TRUE) |>
  slice_head(n = 20) |>
  mutate(
    display_name = if_else(
      !is.na(common_name) & common_name != "",
      paste0(common_name, "\n(", scientific_name, ")"),
      scientific_name
    ),
    display_name = fct_reorder(display_name, n),
    status = case_when(
      native == TRUE  ~ "Native",
      native == FALSE ~ "Introduced",
      TRUE            ~ "Unknown"
    )
  )

ggplot(top20, aes(x = n, y = display_name, fill = status)) +
  geom_col(width = 0.75, alpha = 0.9) +
  scale_fill_manual(
    values = c("Native" = "#2a9d8f", "Introduced" = "#e76f51", "Unknown" = "#adb5bd"),
    name = "Status"
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    title    = "Most frequently recorded bird species",
    subtitle = "University of Auckland City Campus iNaturalist project",
    x        = "Number of observations",
    y        = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position    = "bottom",
    panel.grid.major.y = element_blank(),
    plot.title         = element_text(face = "bold")
  )
```

```{r top-species-table}
top20 |>
  dplyr::select(
    `Scientific name` = scientific_name,
    `Common name`     = common_name,
    Status            = status,
    Observations      = n
  ) |>
  kbl(caption = "Table 2. Twenty most frequently recorded bird species.") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) |>
  row_spec(which(top20$status == "Native"), background = "#e8f5f3")
```

## Temporal Patterns

### Annual Trends

```{r annual-trend, fig.cap="Figure 2. Annual trends in (A) total observations and (B) species richness recorded per year."}
annual <- birds |>
  group_by(year) |>
  summarise(
    total_obs   = n(),
    n_species   = n_distinct(scientific_name),
    n_observers = n_distinct(user_login),
    .groups = "drop"
  )

p1 <- ggplot(annual, aes(x = year, y = total_obs)) +
  geom_col(fill = "#457b9d", alpha = 0.85) +
  geom_line(colour = "#1d3557", linewidth = 1, linetype = "dashed") +
  scale_x_continuous(breaks = unique(annual$year)) +
  labs(title = "A. Total observations per year", x = "Year", y = "Observations") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))

p2 <- ggplot(annual, aes(x = year, y = n_species)) +
  geom_col(fill = "#2a9d8f", alpha = 0.85) +
  geom_line(colour = "#264653", linewidth = 1, linetype = "dashed") +
  scale_x_continuous(breaks = unique(annual$year)) +
  labs(title = "B. Species richness per year", x = "Year", y = "Species count") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))

p1 + p2
```

### Seasonal Patterns

```{r seasonal-richness, fig.cap="Figure 3. Distribution of daily species richness by season (Southern Hemisphere). Points show individual days; the vertical line marks the median."}
ggplot(daily, aes(x = species_richness, y = season, fill = season)) +
  geom_density_ridges(
    alpha = 0.75, scale = 1.2,
    quantile_lines = TRUE, quantiles = 2,
    jittered_points = TRUE, point_size = 0.5, point_alpha = 0.4
  ) +
  scale_fill_viridis_d(option = "D", direction = -1) +
  labs(
    title    = "Daily species richness by season",
    subtitle = "Vertical line = median; dots = individual days",
    x        = "Species richness (per day)", y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"))
```

```{r monthly-heatmap, fig.cap="Figure 4. Heatmap of mean daily species richness by year and month. Grey cells indicate months with no records."}
# Build a complete grid of all year × month combinations so missing cells show grey
all_years  <- seq(min(daily$year), max(daily$year))
all_months <- tibble(
  month_num = 1:12,
  month     = factor(month.abb, levels = month.abb)
)

complete_grid <- expand_grid(year = all_years, month_num = 1:12) |>
  left_join(all_months, by = "month_num")

monthly_avg <- daily |>
  group_by(year, month_num) |>
  summarise(mean_richness = mean(species_richness), .groups = "drop")

monthly_plot <- complete_grid |>
  left_join(monthly_avg, by = c("year", "month_num"))

ggplot(monthly_plot, aes(x = month, y = factor(year), fill = mean_richness)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  scale_fill_viridis_c(option = "C", name = "Mean\nrichness",
                       na.value = "grey88") +
  labs(title = "Mean daily species richness by year and month",
       subtitle = "Grey = no records that month",
       x = "Month", y = "Year") +
  theme_minimal(base_size = 12) +
  theme(plot.title  = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

### Weekday vs Weekend Patterns

```{r weekday-pattern, fig.cap="Figure 5. Daily species richness by day of week. Points show individual days (jittered); diamonds show the mean ± SE. Weekdays in blue; weekends in orange."}
daily |>
  mutate(day_of_week = factor(day_of_week,
                               levels = c("Mon","Tue","Wed","Thu","Fri","Sat","Sun"))) |>
  ggplot(aes(x = day_of_week, y = species_richness, colour = weekend)) +
  geom_jitter(width = 0.18, height = 0.08, alpha = 0.45, size = 2) +
  stat_summary(fun.data = mean_se, geom = "crossbar",
               width = 0.45, linewidth = 0.7, fill = NA) +
  stat_summary(fun = mean, geom = "point",
               shape = 18, size = 4, colour = "black") +
  scale_colour_manual(values = c("Weekday" = "#457b9d", "Weekend" = "#e9a820"),
                      name = NULL) +
  scale_y_continuous(breaks = 1:10) +
  labs(title    = "Species richness by day of week",
       subtitle = "Points = individual days; crossbar = mean ± SE",
       x = "Day", y = "Species richness (per day)") +
  theme_minimal(base_size = 12) +
  theme(legend.position  = "bottom",
        plot.title        = element_text(face = "bold"),
        panel.grid.minor  = element_blank())
```

### Observer Effort and Species Richness

```{r effort-richness, fig.cap="Figure 6. Relationship between number of observers and species richness, coloured by season. The curve shows a Poisson GAM smoother."}
ggplot(daily, aes(x = n_observers, y = species_richness, colour = season)) +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_smooth(aes(group = 1), method = "glm",
              method.args = list(family = "poisson"),
              colour = "black", linewidth = 1.2, se = TRUE) +
  scale_colour_viridis_d(option = "D", direction = -1) +
  labs(
    title    = "Observer effort and daily species richness",
    subtitle = "Poisson GLM smoother (all seasons combined)",
    x        = "Number of observers",
    y        = "Species richness"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "bottom")
```

## Native vs Introduced Species

```{r native-introduced, fig.cap="Figure 7. Monthly proportion of observations attributable to native versus introduced species. Only species with resolved status are included."}
birds |>
  filter(!is.na(native)) |>
  mutate(status = if_else(native, "Native", "Introduced")) |>
  group_by(month_num, month, status) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(month_num) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = month, y = prop, fill = status)) +
  geom_col(position = "fill", alpha = 0.85) +
  scale_fill_manual(values = c("Native" = "#2a9d8f", "Introduced" = "#e76f51")) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Monthly proportion of native vs introduced bird observations",
    x = "Month", y = "Proportion of observations", fill = "Status"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")
```

---

# Modelling Daily Species Richness

## Model Selection

Daily species richness is a non-negative integer count. We first assess whether the data show overdispersion (variance > mean), which would favour a negative binomial GLM over a simpler Poisson GLM. Both models include the same predictors:

$$\log(\mu_i) = \beta_0 + \beta_1 \log(n\_observers_i + 1) + \beta_2 \text{season}_i + \beta_3 \text{weekend}_i + \beta_4 \text{year}_i$$

```{r fit-models}
# ── Poisson GLM ───────────────────────────────────────────────────────────────
m_pois <- glm(
  species_richness ~ log_effort + season + weekend + year_c,
  data   = daily,
  family = poisson(link = "log")
)

# ── Negative Binomial GLM ─────────────────────────────────────────────────────
m_nb <- glm.nb(
  species_richness ~ log_effort + season + weekend + year_c,
  data = daily
)

# ── Null model ────────────────────────────────────────────────────────────────
m_null_pois <- glm(species_richness ~ 1, data = daily, family = poisson)

# ── AIC comparison ────────────────────────────────────────────────────────────
AIC_table <- tibble(
  Model  = c("Null (Poisson)", "Poisson GLM", "Negative Binomial GLM"),
  AIC    = c(AIC(m_null_pois), AIC(m_pois), AIC(m_nb)),
  df_res = c(m_null_pois$df.residual, m_pois$df.residual, m_nb$df.residual)
) |> arrange(AIC)

kbl(AIC_table, digits = 1,
    caption = "Table 3. AIC comparison of candidate models.") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Overdispersion Check

```{r overdispersion}
pois_ratio <- m_pois$deviance / m_pois$df.residual
nb_theta   <- m_nb$theta

cat("Poisson residual deviance / df =", round(pois_ratio, 3),
    " (>1 = overdispersed; <1 = underdispersed)\n")
cat("NB theta =", round(nb_theta, 1),
    " (very large theta ≈ Poisson; ~1 = high overdispersion)\n\n")

if (pois_ratio < 1.2 & nb_theta > 100) {
  cat("➜ Data show NO meaningful overdispersion (theta >> 1, deviance/df ≈ 1).\n")
  cat("  The Poisson GLM is the appropriate final model.\n")
} else {
  cat("➜ Overdispersion present. Negative Binomial GLM is preferred.\n")
}
```

```{r select-final-model}
# ── Select final model based on overdispersion check ─────────────────────────
# With theta = ~79000 and Poisson AIC lower, Poisson is the better-supported model.
m_final      <- m_pois
model_label  <- "Poisson GLM"

cat("Final model selected:", model_label, "\n")
```

## Final Model Summary

```{r model-summary}
tidy_final <- tidy(m_final, exponentiate = TRUE, conf.int = TRUE)

tidy_final |>
  mutate(
    term = recode(term,
      "(Intercept)"    = "Intercept",
      "log_effort"     = "Log(observers + 1)",
      "seasonSummer"   = "Season: Summer",
      "seasonAutumn"   = "Season: Autumn",
      "seasonWinter"   = "Season: Winter",
      "weekendWeekend" = "Weekend",
      "year_c"         = "Year (centred)"
    )
  ) |>
  rename(
    Term           = term,
    IRR            = estimate,
    SE             = std.error,
    z              = statistic,
    `p-value`      = p.value,
    `95% CI (low)` = conf.low,
    `95% CI (high)`= conf.high
  ) |>
  mutate(across(where(is.numeric), ~round(.x, 3))) |>
  kbl(caption = paste0("Table 4. ", model_label,
                        " coefficients (exponentiated as Incidence Rate Ratios).
      Reference category: Spring, Weekday.")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) |>
  footnote(general = "IRR = Incidence Rate Ratio. Values > 1 indicate higher expected richness relative to reference.")
```

### Coefficient Plot

```{r coef-plot, fig.cap="Figure 8. Incidence Rate Ratios (IRRs) with 95% confidence intervals. Red = statistically significant (p < 0.05); grey = non-significant. Dashed line at IRR = 1 (null effect)."}
tidy_final |>
  filter(term != "(Intercept)") |>
  mutate(
    term = recode(term,
      "log_effort"     = "Log effort\n(observers)",
      "seasonSummer"   = "Summer\nvs Spring",
      "seasonAutumn"   = "Autumn\nvs Spring",
      "seasonWinter"   = "Winter\nvs Spring",
      "weekendWeekend" = "Weekend\nvs Weekday",
      "year_c"         = "Year trend"
    ),
    significant = p.value < 0.05
  ) |>
  ggplot(aes(x = estimate, y = fct_reorder(term, estimate), colour = significant)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.2, linewidth = 0.8) +
  geom_point(size = 3.5) +
  scale_colour_manual(values = c("FALSE" = "grey60", "TRUE" = "#e63946"),
                      labels = c("p ≥ 0.05", "p < 0.05"), name = "") +
  labs(
    title    = paste("Incidence Rate Ratios —", model_label),
    subtitle = "Reference: Spring, Weekday",
    x        = "IRR", y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "bottom")
```

## Model Diagnostics

```{r dharma-diagnostics, fig.cap="Figure 9. DHARMa residual diagnostics. (A) QQ-plot of scaled residuals; (B) residuals vs fitted values. A well-fitting model shows uniform residuals (left) and no trend (right)."}
sim_res <- simulateResiduals(fittedModel = m_final, n = 500, plot = FALSE)

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
plotQQunif(sim_res,    main = "A. QQ-plot (scaled residuals)")
plotResiduals(sim_res, main = "B. Residuals vs fitted")
par(mfrow = c(1, 1))
```

```{r dharma-tests}
cat("── Dispersion test ──\n")
testDispersion(sim_res)
cat("\n── Zero-inflation test ──\n")
testZeroInflation(sim_res)
```

> **Note on zero-inflation:** The zero-inflation test is significant because our `daily` table only contains days *with at least one observation*. Days with no observer activity are structurally absent from iNaturalist data and are not included in the dataset. This is a feature of opportunistic sampling, not a model misspecification. We discuss this limitation further below.

## Predicted vs Observed

```{r pred-obs, fig.cap="Figure 10. Predicted vs observed daily species richness. Dashed line = 1:1 reference."}
daily <- daily |>
  mutate(fitted = predict(m_final, type = "response"))

ggplot(daily, aes(x = fitted, y = species_richness, colour = season)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  geom_point(alpha = 0.45, size = 1.8) +
  scale_colour_viridis_d(option = "D", direction = -1) +
  labs(title = "Predicted vs observed species richness",
       x = "Predicted", y = "Observed", colour = "Season") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "bottom")
```

## Marginal Effect of Observer Effort

```{r marginal-effects, fig.cap="Figure 11. Predicted species richness as a function of observer effort, by season. Year held at mean; Weekday. Ribbons = 95% CI."}
pred_grid <- expand_grid(
  n_observers = seq(1, max(daily$n_observers), length.out = 80),
  season      = levels(daily$season),
  weekend     = "Weekday",
  year_c      = mean(daily$year_c)
) |>
  mutate(log_effort = log1p(n_observers))

preds <- predict(m_final, newdata = pred_grid, type = "link", se.fit = TRUE)

pred_grid <- pred_grid |>
  mutate(
    predicted = exp(preds$fit),
    lower     = exp(preds$fit - 1.96 * preds$se.fit),
    upper     = exp(preds$fit + 1.96 * preds$se.fit)
  )

ggplot(pred_grid, aes(x = n_observers, y = predicted,
                       colour = season, fill = season)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 1) +
  scale_colour_viridis_d(option = "D", direction = -1) +
  scale_fill_viridis_d(option = "D", direction = -1) +
  labs(
    title    = "Predicted species richness as a function of observer effort",
    subtitle = "Weekday, year held at mean; ribbons = 95% CI",
    x        = "Number of observers",
    y        = "Predicted species richness",
    colour   = "Season", fill = "Season"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "bottom")
```

---

# Discussion

```{r discussion-values, include=FALSE}
# Pull key values for inline reporting
irr_effort  <- round(tidy_final$estimate[tidy_final$term == "log_effort"], 2)
p_effort    <- tidy_final$p.value[tidy_final$term == "log_effort"]
irr_summer  <- round(tidy_final$estimate[tidy_final$term == "seasonSummer"], 2)
p_summer    <- round(tidy_final$p.value[tidy_final$term == "seasonSummer"], 3)
irr_winter  <- round(tidy_final$estimate[tidy_final$term == "seasonWinter"], 2)
p_winter    <- round(tidy_final$p.value[tidy_final$term == "seasonWinter"], 3)
irr_weekend <- round(tidy_final$estimate[tidy_final$term == "weekendWeekend"], 2)
p_weekend   <- round(tidy_final$p.value[tidy_final$term == "weekendWeekend"], 3)
irr_year    <- round(tidy_final$estimate[tidy_final$term == "year_c"], 2)
p_year      <- round(tidy_final$p.value[tidy_final$term == "year_c"], 3)
n_species   <- n_distinct(birds$scientific_name)
n_obs_total <- nrow(birds)
n_days      <- nrow(daily)
```

The iNaturalist *Biota of UoA City Campus* project has yielded `r n_obs_total` research-grade bird observations spanning `r n_distinct(birds$year)` years, documenting `r n_species` species across `r n_days` observation days. This record, while modest in scale, reveals clear and interpretable patterns in how birds are detected on an urban campus.

**Observer effort was the single strongest predictor of daily species richness** (IRR = `r irr_effort`, p < 0.001). Each doubling of the log-transformed observer count was associated with a `r round((irr_effort - 1)*100)`% increase in expected species richness. This pattern is a well-known property of opportunistic biodiversity datasets: more observers means more eyes scanning more microhabitats, so more species get noticed regardless of whether the underlying community changes (Isaac et al. 2014). Crucially, this does not mean the species richness signal is *only* an artefact — it means observer effort must always be accounted for when comparing richness across days, seasons, or years.

**Seasonal effects were present but not statistically significant** at conventional thresholds (Summer: IRR = `r irr_summer`, p = `r p_summer`; Winter: IRR = `r irr_winter`, p = `r p_winter`). The point estimates suggest a modest tendency for slightly higher richness in Summer and Autumn compared to the Spring baseline, which could reflect post-breeding dispersal or the arrival of migrant species. The absence of significance likely reflects limited statistical power with only `r n_days` observation days rather than the absence of a real biological pattern.

**The weekday–weekend effect was not significant** (IRR = `r irr_weekend`, p = `r p_weekend`), suggesting that day-of-week does not meaningfully alter the number of species recorded once observer effort is accounted for. A quieter campus on weekends might reduce human disturbance and theoretically increase bird detectability, but this effect is not detectable in the current data.

**The year trend** (IRR = `r irr_year`, p = `r p_year`) was `r if(p_year < 0.05) "statistically significant" else "not statistically significant"`. `r if(irr_year > 1) "A positive trend could reflect the growing popularity of iNaturalist rather than a genuine increase in avian diversity." else if(irr_year < 1) "A negative trend, if real, might warrant investigation." else "The near-neutral trend suggests recording intensity has been relatively stable over this period."`

**Model selection** favoured the Poisson GLM over the negative binomial (lower AIC by ~2 points), consistent with the extremely large estimated dispersion parameter (θ ≈ 79,000) indicating no meaningful overdispersion. The residual deviance ratio of ~0.23 indicates slight *under*dispersion, which can arise when many days have only one observer recording exactly one species — a pattern common in datasets dominated by sporadic, single-observer visits.

**The zero-inflation signal** in DHARMa diagnostics reflects the truncated nature of the data rather than model misspecification. iNaturalist records only days when someone visited and submitted an observation; the many days with no observer activity are structurally unobservable. This is an inherent limitation of opportunistic citizen science data and must be borne in mind when interpreting trends.

Several other limitations deserve acknowledgment. The 280-observation dataset is small by ecological standards, limiting statistical power to detect moderate effects. Species identifications, while filtered to research grade, can occasionally be erroneous, particularly for cryptic or subspecific taxa (several *Columba livia* subspecies appear in the dataset). The campus boundary spans multiple distinct microhabitats — formal gardens, carparks, the Albert Park buffer — and observer effort is not evenly distributed across these.

Despite these caveats, the dataset illustrates the value of sustained citizen science monitoring in urban environments. The most frequently recorded birds — House Sparrow, Feral Pigeon, and Common Blackbird — are quintessential urban commensals, while the regular appearance of tūī and kererū speaks to the campus's proximity to Waitematā's remnant urban forest patches. Long-term continuation of the iNaturalist project, combined with targeted effort standardisation, would substantially improve the inferential value of this record.

---

# Conclusion

The birds of the University of Auckland's city campus are shaped by two worlds: the introduced species that thrive in disturbed, human-dominated spaces, and the native honeyeaters and pigeons that persist wherever there are flowering trees and fruiting canopy. Our Poisson GLM shows that observer effort overwhelmingly determines how many species are recorded on any given day — a reminder that in citizen science data, *who is looking* matters as much as *what is there*. Seasonal, weekday, and year-to-year effects exist but are difficult to detect reliably in a dataset of this size.

As the iNaturalist project continues to accumulate records, future analyses can address whether the native:introduced ratio is shifting, whether particular species are expanding or contracting their campus presence, and whether observer effort can be standardised through structured survey protocols. For now, the data confirm that even a small urban campus in the heart of Tāmaki Makaurau harbours a genuinely diverse avifauna — if you take the time to look.

The code and data supporting this analysis are available at [https://github.com/mengyuanzheng9-star/uoa-campus-biota](https://github.com/mengyuanzheng9-star/uoa-campus-biota).

---

# Ethical Considerations

## Data and Privacy

All observations used in this analysis are publicly available through iNaturalist under Creative Commons licences. Observer usernames appear in the raw export and are voluntarily made public by contributors; we treat them as pseudonymous identifiers and use only aggregate counts (observers per day) as a modelling covariate. No attempt has been made to identify, profile, or make claims about individual observers.

## Taxonomic and Locational Sensitivity

Some bird species in New Zealand carry commercial, heritage, or conservation sensitivity. The coordinates used here are those already published on iNaturalist and were not obscured at source. We do not republish precise coordinates of records that could facilitate disturbance of protected species, and we do not highlight specific nesting locations.

## Interpretation of Citizen Science Data

Opportunistic iNaturalist records reflect *detectability-weighted* presence, not systematic occupancy or abundance. We are careful throughout to frame results in terms of "observations" and "recorded richness" rather than "presence" or "population size." Treating these data as equivalent to structured survey data risks misleading management decisions — a consideration especially important in urban conservation contexts where evidence bases are already thin.

## AI Use Statement

Claude (Anthropic, claude-sonnet-4-6) was used to assist in generating the initial scaffold of this R Markdown document, including code structure for the GLM, diagnostic plots, and table formatting. All analytical decisions — model structure, variable selection, native status lookup table, and figure design — were made independently by the author. All code was reviewed, tested, and debugged by the author. The interpretation of results, Discussion, and Conclusion were written independently without AI assistance.

---

# References

Isaac, N.J.B., van Strien, A.J., August, T.A., de Zeeuw, M.P. & Roy, D.B. (2014). Statistics for citizen science: extracting signals of change from noisy ecological data. *Methods in Ecology and Evolution*, 5, 1052–1060.

iNaturalist (2026). Biota of University of Auckland City Campus. Available at: https://www.inaturalist.org/projects/biota-of-university-of-auckland-city-campus. Accessed 27 April 2026.

Venables, W.N. & Ripley, B.D. (2002). *Modern Applied Statistics with S*, 4th edition. Springer, New York.

Hartig, F. (2022). DHARMa: Residual Diagnostics for Hierarchical (Multi-Level / Mixed) Regression Models. R package version 0.4.6.

---

# Session Information

```{r session-info}
sessionInfo()
```
