---
title: "The Birds We Notice: Observer Effort, Citizen Science Bias, and Avian Diversity on the University of Auckland City Campus"
author: "Mengyuan Zheng"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    fig_caption: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.align  = "center",
  fig.width  = 8,
  fig.height = 5,
  dpi        = 150
)
```

---

> **GitHub Repository:** [https://github.com/mengyuanzheng9-star/uoa-campus-biota](https://github.com/mengyuanzheng9-star/uoa-campus-biota)

---

# Introduction

Imagine a restaurant review platform where the top five most active reviewers have written a quarter of all reviews. Would you trust the ranking? Would "most reviewed dish" reliably mean "best dish" — or would it simply mean "the dish those five reviewers happened to order most often"?

This is not a hypothetical problem. It is the situation facing anyone who opens the iNaturalist dataset for the University of Auckland's city campus. Among 138 contributors who have submitted research-grade bird observations since 2016, the five most active observers account for 71 records — approximately 25% of the entire dataset. Before we can say anything meaningful about which birds inhabit this campus, we need to ask: does that concentration of effort distort the picture?

This analysis pursues that question in three stages. First, we describe the campus bird record and examine the distribution of observer effort. Second, we test whether the ranking of bird groups is robust to the removal of the most active contributors — a sensitivity analysis analogous to asking whether the restaurant ranking holds once the power reviewers are excluded. Third, having established the reliability of the main pattern, we use statistical modelling to ask what actually drives variation in daily species richness: is it the season, the day of the week, the year, or simply how many observers were active on a given day?

The iNaturalist [Biota of University of Auckland City Campus](https://www.inaturalist.org/projects/biota-of-university-of-auckland-city-campus) project provides the data. The findings speak not only to campus ecology, but to the broader challenge of interpreting opportunistic citizen science records honestly.

---

# Methods

## Data Acquisition

Observation data were exported from the iNaturalist project *Biota of University of Auckland City Campus* on **27 April 2026**, filtering to:

- **Taxon:** Aves (birds)
- **Quality grade:** Research grade
- **Date range:** All available records to export date

The exported CSV (`observations-719197.csv`) was saved in the project repository alongside this script. No location data were obscured in the downloaded export.

## Data Wrangling

```{r libraries}
# ── Core data manipulation ──────────────────────────────────────────────────
library(tidyverse)
library(lubridate)
library(janitor)

# ── Modelling ───────────────────────────────────────────────────────────────
library(MASS)          # glm.nb; masks dplyr::select — restored below
select <- dplyr::select
library(lme4)          # glmer for mixed-effects model
library(broom)
library(broom.mixed)   # tidy() for lme4 objects
library(DHARMa)

# ── Visualisation ───────────────────────────────────────────────────────────
library(patchwork)
library(ggridges)
library(viridis)
library(knitr)
library(kableExtra)
library(vegan)        # species accumulation / rarefaction
```

```{r native-lookup}
# ── Manual native/introduced status lookup ───────────────────────────────────
# Based on NZ Birds Online and iNaturalist taxonomy.
# TRUE = native to NZ; FALSE = introduced/naturalised.
# Includes parent species AND subspecies names that appear in this dataset.
native_status <- tribble(
  ~scientific_name,                                ~native,
  # ── Native species (parent + subspecies) ────────────────────────────────
  "Prosthemadera novaeseelandiae",                 TRUE,
  "Prosthemadera novaeseelandiae novaeseelandiae", TRUE,
  "Hemiphaga novaeseelandiae",                     TRUE,
  "Hemiphaga novaeseelandiae novaeseelandiae",     TRUE,
  "Rhipidura fuliginosa",                          TRUE,
  "Rhipidura fuliginosa fuliginosa",               TRUE,
  "Rhipidura fuliginosa placabilis",               TRUE,   # North Island Fantail
  "Rhipidura fuliginosa melanura",                 TRUE,   # South Island Fantail
  "Todiramphus sanctus",                           TRUE,
  "Todiramphus sanctus vagans",                    TRUE,
  "Halcyon sancta",                                TRUE,
  "Zosterops lateralis",                           TRUE,   # Silvereye (native coloniser)
  "Zosterops lateralis lateralis",                 TRUE,
  "Anthornis melanura",                            TRUE,
  "Anthornis melanura melanura",                   TRUE,
  "Petroica macrocephala",                         TRUE,
  "Gerygone igata",                                TRUE,
  "Mohoua ochrocephala",                           TRUE,
  "Anas platyrhynchos superciliosa",               TRUE,   # Grey Duck
  "Ardea novaehollandiae",                         TRUE,   # White-faced Heron
  "Porphyrio melanotus",                           TRUE,   # Pūkeko
  "Porphyrio melanotus melanotus",                 TRUE,
  "Haematopus ostralegus",                         TRUE,
  "Anthus novaeseelandiae",                        TRUE,   # NZ Pipit
  "Anthus novaeseelandiae novaeseelandiae",         TRUE,
  # ── Introduced species (parent + subspecies) ────────────────────────────
  "Passer domesticus",                             FALSE,
  "Passer domesticus domesticus",                  FALSE,
  "Columba livia",                                 FALSE,
  "Columba livia domestica",                       FALSE,
  "Turdus merula",                                 FALSE,
  "Turdus merula merula",                          FALSE,
  "Platycercus eximius",                           FALSE,
  "Platycercus eximius eximius",                   FALSE,
  "Gymnorhina tibicen",                            FALSE,
  "Gymnorhina tibicen hypoleuca",                  FALSE,
  "Gymnorhina tibicen tibicen",                    FALSE,
  "Sturnus vulgaris",                              FALSE,
  "Carduelis carduelis",                           FALSE,
  "Carduelis carduelis britannica",                FALSE,
  "Carduelis chloris",                             FALSE,
  "Fringilla coelebs",                             FALSE,
  "Fringilla coelebs gengleri",                    FALSE,
  "Anas platyrhynchos",                            FALSE,  # Mallard
  "Anas platyrhynchos platyrhynchos",              FALSE,
  "Streptopelia chinensis",                        FALSE,
  "Streptopelia chinensis tigrina",                FALSE,
  "Streptopelia roseogrisea",                      FALSE,
  "Acridotheres tristis",                          FALSE,
  "Acridotheres tristis tristis",                  FALSE,
  "Corvus frugilegus",                             FALSE,
  "Turdus philomelos",                             FALSE,
  "Turdus philomelos clarkei",                     FALSE,
  "Hirundo neoxena",                               TRUE,   # Welcome Swallow (native coloniser)
  "Hirundo neoxena neoxena",                       TRUE,
  "Hirundo tahitica",                              FALSE,
  "Ardea alba",                                    FALSE   # Great Egret (vagrant)
)
```

```{r load-data}
# ── Load raw data ─────────────────────────────────────────────────────────────
raw_path <- "observations-719197.csv"

if (file.exists(raw_path)) {
  raw <- read_csv(raw_path, show_col_types = FALSE)
  cat("✓ Real data loaded:", nrow(raw), "rows,", ncol(raw), "columns\n")
} else {
  stop("Data file not found: ", raw_path,
       "\nPlease place observations-719197.csv in the same folder as this Rmd.")
}
```

```{r clean-data}
# ── Standardise column names ──────────────────────────────────────────────────
raw_clean <- raw |> clean_names()

# Confirm Aves filter if column present
if ("iconic_taxon_name" %in% names(raw_clean)) {
  non_bird <- raw_clean |> filter(iconic_taxon_name != "Aves")
  if (nrow(non_bird) > 0) {
    cat("Warning:", nrow(non_bird), "non-Aves rows removed.\n")
    raw_clean <- raw_clean |> filter(iconic_taxon_name == "Aves")
  }
}

# ── Resolve column names robustly ─────────────────────────────────────────────
date_col     <- intersect(c("observed_on", "observed_on_string"), names(raw_clean))[1]
observer_col <- intersect(c("user_login", "user_id"), names(raw_clean))[1]

# ── Build birds tibble ────────────────────────────────────────────────────────
birds <- raw_clean |>
  rename(observed_on_col = all_of(date_col),
         user_login      = all_of(observer_col)) |>
  mutate(
    date        = as.Date(substr(as.character(observed_on_col), 1, 10)),
    year        = year(date),
    month       = month(date, label = TRUE, abbr = TRUE),
    month_num   = month(date),
    day_of_week = wday(date, label = TRUE, abbr = TRUE, week_start = 1),
    season      = case_when(
      month_num %in% c(12, 1, 2)  ~ "Summer",
      month_num %in% c(3, 4, 5)   ~ "Autumn",
      month_num %in% c(6, 7, 8)   ~ "Winter",
      month_num %in% c(9, 10, 11) ~ "Spring"
    ),
    season  = factor(season, levels = c("Spring", "Summer", "Autumn", "Winter")),
    weekend = if_else(wday(date, week_start = 1) >= 6, "Weekend", "Weekday")
  ) |>
  left_join(native_status, by = "scientific_name") |>
  filter(!is.na(date), year >= 2015) |>
  distinct()

cat("Total bird observations:", nrow(birds), "\n")
cat("Date range:", as.character(min(birds$date)), "to", as.character(max(birds$date)), "\n")
cat("Unique species:", n_distinct(birds$scientific_name), "\n")
cat("Unique observers:", n_distinct(birds$user_login), "\n")
cat("Native status resolved for",
    sum(!is.na(birds$native)), "/", nrow(birds), "observations\n")
```

## Derived Variables

For daily-level modelling, we aggregate to each calendar day with at least one record:

| Variable | Description |
|---|---|
| `species_richness` | Number of distinct species observed |
| `n_observers` | Number of unique observers (effort proxy) |
| `log_effort` | log(n_observers + 1), stabilises variance |
| `season` | Southern Hemisphere astronomical season |
| `weekend` | Weekend vs Weekday |
| `year_c` | Calendar year centred at first year |

```{r daily-summary}
daily <- birds |>
  group_by(date, year, month, month_num, season, weekend, day_of_week) |>
  summarise(
    species_richness = n_distinct(scientific_name),
    n_obs            = n(),
    n_observers      = n_distinct(user_login),
    .groups = "drop"
  ) |>
  mutate(
    year_c     = year - min(year),
    log_effort = log1p(n_observers)
  )

cat("Days with at least one bird observation:", nrow(daily), "\n")
cat("\nSpecies richness per day:\n")
print(summary(daily$species_richness))
```

---

# Results

## Dataset Overview

```{r overview-table}
overview <- tibble(
  Metric = c("Total observations", "Research-grade records",
             "Unique bird species", "Unique observers",
             "First observation", "Last observation",
             "Days with ≥1 record"),
  Value  = c(
    format(nrow(birds), big.mark = ","),
    format(sum(birds$quality_grade == "research", na.rm = TRUE), big.mark = ","),
    n_distinct(birds$scientific_name),
    n_distinct(birds$user_login),
    as.character(min(birds$date)),
    as.character(max(birds$date)),
    nrow(daily)
  )
)

kbl(overview, caption = "Table 1. Dataset overview — Biota of UoA City Campus (Aves)") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

## The Campus Bird Ranking

```{r bird-group-lookup}
# ── Collapse species and subspecies into readable bird groups ─────────────────
# This mirrors the data-cleaning approach described in the article:
# subspecies are merged with their parent species for public-facing clarity.
bird_group_lookup <- tribble(
  ~scientific_name,                                ~bird_group,
  "Columba livia",                                 "Rock/Feral Pigeon",
  "Columba livia domestica",                       "Rock/Feral Pigeon",
  "Passer domesticus",                             "House Sparrow",
  "Passer domesticus domesticus",                  "House Sparrow",
  "Turdus merula",                                 "Eurasian Blackbird",
  "Turdus merula merula",                          "Eurasian Blackbird",
  "Prosthemadera novaeseelandiae",                 "Tūī",
  "Prosthemadera novaeseelandiae novaeseelandiae", "Tūī",
  "Platycercus eximius",                           "Eastern Rosella",
  "Platycercus eximius eximius",                   "Eastern Rosella",
  "Rhipidura fuliginosa",                          "NZ Fantail / Pīwakawaka",
  "Rhipidura fuliginosa fuliginosa",               "NZ Fantail / Pīwakawaka",
  "Rhipidura fuliginosa placabilis",               "NZ Fantail / Pīwakawaka",
  "Rhipidura fuliginosa melanura",                 "NZ Fantail / Pīwakawaka",
  "Hemiphaga novaeseelandiae",                     "Kererū",
  "Hemiphaga novaeseelandiae novaeseelandiae",     "Kererū",
  "Gymnorhina tibicen",                            "Australian Magpie",
  "Gymnorhina tibicen hypoleuca",                  "Australian Magpie",
  "Gymnorhina tibicen tibicen",                    "Australian Magpie",
  "Zosterops lateralis",                           "Silvereye / Tauhou",
  "Zosterops lateralis lateralis",                 "Silvereye / Tauhou",
  "Todiramphus sanctus",                           "Sacred Kingfisher",
  "Todiramphus sanctus vagans",                    "Sacred Kingfisher",
  "Halcyon sancta",                                "Sacred Kingfisher",
  "Carduelis carduelis",                           "European Goldfinch",
  "Carduelis carduelis britannica",                "European Goldfinch",
  "Hirundo neoxena",                               "Welcome Swallow / Warou",
  "Hirundo neoxena neoxena",                       "Welcome Swallow / Warou",
  "Sturnus vulgaris",                              "Common Starling",
  "Acridotheres tristis",                          "Common Myna",
  "Acridotheres tristis tristis",                  "Common Myna",
  "Turdus philomelos",                             "Song Thrush",
  "Turdus philomelos clarkei",                     "Song Thrush",
  "Fringilla coelebs",                             "Chaffinch",
  "Fringilla coelebs gengleri",                    "Chaffinch",
  "Carduelis chloris",                             "European Greenfinch",
  "Anas platyrhynchos",                            "Mallard",
  "Anas platyrhynchos platyrhynchos",              "Mallard",
  "Anas platyrhynchos superciliosa",               "Grey Duck",
  "Gerygone igata",                                "Grey Warbler / Riroriro",
  "Corvus frugilegus",                             "Rook",
  "Streptopelia chinensis",                        "Spotted Dove",
  "Streptopelia chinensis tigrina",                "Spotted Dove",
  "Ardea novaehollandiae",                         "White-faced Heron",
  "Ardea alba",                                    "Great Egret",
  "Anthornis melanura",                            "Bellbird / Korimako",
  "Anthornis melanura melanura",                   "Bellbird / Korimako",
  "Hirundo tahitica",                              "Pacific Swallow",
  "Streptopelia roseogrisea",                      "African Collared Dove",
  "Porphyrio melanotus",                           "Pūkeko",
  "Porphyrio melanotus melanotus",                 "Pūkeko",
  "Haematopus ostralegus",                         "Oystercatcher",
  "Anthus novaeseelandiae",                        "NZ Pipit",
  "Anthus novaeseelandiae novaeseelandiae",         "NZ Pipit",
  "Petroica macrocephala",                         "Tomtit / Miromiro",
  "Mohoua ochrocephala",                           "Yellowhead / Mohua"
)

# ── Join bird_group to birds tibble ──────────────────────────────────────────
birds <- birds |>
  left_join(bird_group_lookup, by = "scientific_name") |>
  mutate(bird_group = if_else(is.na(bird_group), scientific_name, bird_group))
```

```{r top-groups, fig.width=9, fig.height=7, fig.cap="Figure 1. Most frequently recorded bird groups on the UoA city campus. Native NZ species shown in teal; introduced species in salmon."}
# ── Compute group-level counts ────────────────────────────────────────────────
group_counts <- birds |>
  group_by(bird_group) |>
  summarise(
    n      = n(),
    native = first(native),
    .groups = "drop"
  ) |>
  mutate(
    status       = case_when(
      native == TRUE  ~ "Native",
      native == FALSE ~ "Introduced",
      TRUE            ~ "Unknown"
    ),
    bird_group = fct_reorder(bird_group, n)
  ) |>
  arrange(desc(n))

# Store ranking for sensitivity analysis
ranking_all <- group_counts |>
  arrange(desc(n)) |>
  mutate(rank_all = row_number())

ggplot(group_counts, aes(x = n, y = bird_group, fill = status)) +
  geom_col(width = 0.75, alpha = 0.9) +
  scale_fill_manual(
    values = c("Native" = "#2a9d8f", "Introduced" = "#e76f51", "Unknown" = "#adb5bd"),
    name = "Status"
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    title    = "Most frequently recorded bird groups",
    subtitle = "University of Auckland City Campus iNaturalist project",
    x        = "Number of observations",
    y        = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position    = "bottom",
    panel.grid.major.y = element_blank(),
    plot.title         = element_text(face = "bold")
  )
```

## Who Is Contributing the Data?

```{r observer-effort, fig.cap="Figure 2. Distribution of records across 138 observers. The top five most active observers (red) together account for approximately 25% of all records."}
# ── Per-observer record counts ────────────────────────────────────────────────
observer_counts <- birds |>
  count(user_login, sort = TRUE) |>
  mutate(
    rank     = row_number(),
    top5     = rank <= 5,
    observer = if_else(top5, paste0("Top-", rank), "Other observers")
  )

top5_users    <- observer_counts |> filter(top5) |> pull(user_login)
top5_n        <- observer_counts |> filter(top5) |> pull(n) |> sum()
top5_pct      <- round(100 * top5_n / nrow(birds), 1)
total_obs_n   <- nrow(birds)
total_obs_pct <- 100 - top5_pct

cat("Top 5 observers contributed:", top5_n, "records (", top5_pct, "% of total)\n")
cat("Remaining", n_distinct(birds$user_login) - 5,
    "observers contributed:", total_obs_n - top5_n,
    "records (", total_obs_pct, "%)\n")

# ── Lollipop chart of observer contributions ─────────────────────────────────
observer_counts |>
  mutate(label = if_else(top5, paste0("Observer ", rank), NA_character_)) |>
  ggplot(aes(x = rank, y = n, colour = top5)) +
  geom_segment(aes(xend = rank, yend = 0), linewidth = 0.6, alpha = 0.7) +
  geom_point(size = 2.5) +
  annotate("rect", xmin = 0.5, xmax = 5.5, ymin = 0, ymax = Inf,
           alpha = 0.08, fill = "#e63946") +
  annotate("text", x = 3, y = max(observer_counts$n) * 0.92,
           label = paste0("Top 5:\n", top5_pct, "% of all records"),
           colour = "#e63946", fontface = "bold", size = 3.5) +
  scale_colour_manual(values = c("FALSE" = "grey60", "TRUE" = "#e63946"),
                      guide = "none") +
  scale_x_continuous(breaks = c(1, 5, seq(20, nrow(observer_counts), 20))) +
  labs(
    title    = "Distribution of records across observers",
    subtitle = paste0(n_distinct(birds$user_login),
                      " observers; top 5 highlighted in red"),
    x        = "Observer rank (most to least active)",
    y        = "Number of records"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

## Sensitivity Analysis: Does the Ranking Hold Without the Top Five?

```{r sensitivity-analysis, fig.width=11, fig.height=8, fig.cap="Figure 3. Bird group ranking with all records (left) versus after removing the top five most active observers (right). Lines connect the same group across both rankings; groups that changed rank are highlighted in orange, with rank numbers shown at each end. The top five groups held their positions exactly."}
# ── Recompute ranking after removing top-5 observers ─────────────────────────
birds_no_top5 <- birds |> filter(!user_login %in% top5_users)
ranking_no_top5 <- birds_no_top5 |>
  count(bird_group, sort = TRUE) |>
  mutate(rank_no_top5 = row_number()) |>
  rename(n_no_top5 = n)

# ── Join both rankings ────────────────────────────────────────────────────────
rank_comparison <- ranking_all |>
  select(bird_group, n_all = n, rank_all, status) |>
  inner_join(ranking_no_top5, by = "bird_group") |>
  mutate(
    rank_changed = rank_all != rank_no_top5,
    highlight    = if_else(rank_changed, "#e9a820", "grey70")
  )

# ── Print rank changes ────────────────────────────────────────────────────────
cat("Bird groups that changed rank after removing top-5 observers:\n")
rank_comparison |>
  filter(rank_changed) |>
  select(bird_group, rank_all, rank_no_top5) |>
  mutate(direction = if_else(rank_no_top5 < rank_all, "↑ moved up", "↓ moved down")) |>
  print()

# ── Bump chart ────────────────────────────────────────────────────────────────
rank_long <- rank_comparison |>
  select(bird_group, rank_all, rank_no_top5, rank_changed, status) |>
  pivot_longer(cols = c(rank_all, rank_no_top5),
               names_to  = "dataset",
               values_to = "rank") |>
  mutate(
    dataset = recode(dataset,
                     "rank_all"      = "All records",
                     "rank_no_top5"  = "Top-5 removed"),
    dataset = factor(dataset, levels = c("All records", "Top-5 removed"))
  )

# Only plot top 15 for clarity
top15_groups <- rank_comparison |>
  filter(rank_all <= 15) |>
  pull(bird_group)

rank_long_top15 <- rank_long |> filter(bird_group %in% top15_groups)

# ── Rank label data for left and right endpoints ──────────────────────────────
rank_labels_left <- rank_long_top15 |>
  filter(dataset == "All records") |>
  mutate(label = as.character(rank))

rank_labels_right <- rank_long_top15 |>
  filter(dataset == "Top-5 removed") |>
  mutate(label = as.character(rank))

ggplot(rank_long_top15,
       aes(x = dataset, y = rank, group = bird_group,
           colour = rank_changed)) +
  # ── Grey stable lines first (background) ─────────────────────────────────
  geom_line(data = rank_long_top15 |> filter(!rank_changed),
            linewidth = 0.6, alpha = 0.35) +
  geom_point(data = rank_long_top15 |> filter(!rank_changed),
             size = 2, alpha = 0.4) +
  # ── Orange changed lines on top (foreground) ─────────────────────────────
  geom_line(data = rank_long_top15 |> filter(rank_changed),
            linewidth = 2.0, alpha = 0.95) +
  geom_point(data = rank_long_top15 |> filter(rank_changed),
             size = 3.5, alpha = 1) +
  # ── Bird group name labels ────────────────────────────────────────────────
  geom_text(data = rank_long_top15 |> filter(dataset == "All records"),
            aes(label = bird_group),
            hjust = 1.08, size = 2.8,
            fontface = if_else(
              rank_long_top15 |>
                filter(dataset == "All records") |>
                pull(rank_changed),
              "bold.italic", "italic"
            )) +
  geom_text(data = rank_long_top15 |> filter(dataset == "Top-5 removed"),
            aes(label = bird_group),
            hjust = -0.08, size = 2.8,
            fontface = if_else(
              rank_long_top15 |>
                filter(dataset == "Top-5 removed") |>
                pull(rank_changed),
              "bold.italic", "italic"
            )) +
  # ── Annotation: stable top 5 ─────────────────────────────────────────────
  annotate("rect", xmin = 0.52, xmax = 1.48,
           ymin = 0.5, ymax = 5.5,
           alpha = 0.06, fill = "#2a9d8f", colour = "#2a9d8f",
           linewidth = 0.4, linetype = "dashed") +
  annotate("text", x = 1, y = 0.2,
           label = "Top 5 unchanged",
           colour = "#2a9d8f", size = 3, fontface = "bold") +
  scale_y_reverse(breaks = 1:15) +
  scale_colour_manual(values = c("FALSE" = "grey70", "TRUE" = "#e9a820"),
                      labels = c("Rank unchanged", "Rank changed"),
                      name   = "") +
  expand_limits(x = c(-0.2, 2.8)) +
  labs(
    title    = "Sensitivity analysis: does removing the top five observers change the ranking?",
    subtitle = "Top 15 groups shown. Orange = rank changed; grey = rank stable.",
    x        = NULL,
    y        = "Rank (1 = most recorded)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(size = 10, colour = "grey40"),
    legend.position  = "bottom",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text.x      = element_text(size = 12, face = "bold")
  )
```

## Temporal Patterns and the Semester Effect

```{r monthly-effort, fig.cap="Figure 4. Monthly total observations (bars) and number of active observers (line) across all years combined. March stands out in both metrics, suggesting a semester-driven surge in recording activity rather than a genuine seasonal peak in bird diversity."}
# ── Monthly observations and active observers ─────────────────────────────────
monthly_summary <- birds |>
  group_by(month_num, month) |>
  summarise(
    total_obs    = n(),
    n_observers  = n_distinct(user_login),
    .groups = "drop"
  ) |>
  mutate(obs_per_observer = total_obs / n_observers)

# Dual-axis plot: bars for observations, line for observer count
obs_scale <- max(monthly_summary$total_obs)
obs_line  <- max(monthly_summary$n_observers)
scale_factor <- obs_scale / obs_line

ggplot(monthly_summary, aes(x = month)) +
  geom_col(aes(y = total_obs), fill = "#457b9d", alpha = 0.75, width = 0.6) +
  geom_line(aes(y = n_observers * scale_factor, group = 1),
            colour = "#e63946", linewidth = 1.2) +
  geom_point(aes(y = n_observers * scale_factor),
             colour = "#e63946", size = 3) +
  scale_y_continuous(
    name = "Total observations (bars)",
    sec.axis = sec_axis(~ . / scale_factor,
                        name = "Active observers (red line)")
  ) +
  labs(
    title    = "Monthly observations and active observers",
    subtitle = "All years combined (2016–2026)",
    x        = "Month"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title  = element_text(face = "bold"),
    axis.title.y.right = element_text(colour = "#e63946"),
    axis.text.y.right  = element_text(colour = "#e63946")
  )
```

```{r annual-trend, fig.cap="Figure 5. Annual trends in total observations and unique observers. The parallel growth in observers and records suggests that increased recording activity, rather than a genuine increase in bird diversity, drives the recent upward trend."}
annual <- birds |>
  group_by(year) |>
  summarise(
    total_obs   = n(),
    n_species   = n_distinct(scientific_name),
    n_observers = n_distinct(user_login),
    .groups = "drop"
  )

p1 <- ggplot(annual, aes(x = year, y = total_obs)) +
  geom_col(fill = "#457b9d", alpha = 0.85) +
  geom_line(colour = "#1d3557", linewidth = 1, linetype = "dashed") +
  scale_x_continuous(breaks = unique(annual$year)) +
  labs(title = "A. Total observations per year", x = "Year", y = "Observations") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))

p2 <- ggplot(annual, aes(x = year, y = n_observers)) +
  geom_col(fill = "#e63946", alpha = 0.75) +
  geom_line(colour = "#9d0208", linewidth = 1, linetype = "dashed") +
  scale_x_continuous(breaks = unique(annual$year)) +
  labs(title = "B. Active observers per year", x = "Year", y = "Observers") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))

p1 + p2
```

---

# Modelling Daily Species Richness

Having established that the main bird ranking is robust to observer identity, we now ask a different question: what drives variation in *how many species* are recorded on any given day? Candidate predictors include observer effort (how many people were active), season, day of the week, and year.

## Model Specification

Daily species richness is a non-negative integer count. We fit a family of models with the same predictors:

$$\log(\mu_i) = \beta_0 + \beta_1 \log(n\_observers_i + 1) + \beta_2 \text{season}_i + \beta_3 \text{weekend}_i + \beta_4 \text{year}_i$$

We compare three fixed-effects models — Null (intercept only), Poisson GLM, and Negative Binomial GLM — and then extend the best-fitting fixed-effects model to include observer identity as a random effect (GLMM), acknowledging that individual observers may have systematically different recording tendencies.

## Fixed-Effects Model Comparison

```{r species-accumulation-curve}
# ── Species accumulation curve: the mechanical baseline ───────────────────────
# Recorded richness is, by construction, a function of how much sampling occurred.
# Before interpreting the effort effect ecologically, we make this baseline explicit.
comm <- birds |>
  count(date, bird_group) |>
  pivot_wider(names_from = bird_group, values_from = n, values_fill = 0) |>
  column_to_rownames("date")

set.seed(738)
accum <- specaccum(comm, method = "random", permutations = 200)

accum_df <- tibble(
  effort   = accum$sites,
  richness = accum$richness,
  sd       = accum$sd
)

ggplot(accum_df, aes(effort, richness)) +
  geom_ribbon(aes(ymin = richness - 1.96 * sd, ymax = richness + 1.96 * sd),
              alpha = 0.15) +
  geom_line(linewidth = 1) +
  labs(
    x = "Sampling effort (number of recording days)",
    y = "Expected cumulative bird groups",
    title = "Species accumulation curve",
    subtitle = "Recorded richness rises with effort and has not yet plateaued"
  ) +
  theme_minimal(base_size = 12)
```

```{r fit-fixed-models}
# ── Poisson GLM ───────────────────────────────────────────────────────────────
m_pois <- glm(
  species_richness ~ log_effort + season + weekend + year_c,
  data   = daily,
  family = poisson(link = "log")
)

# ── Negative Binomial GLM ─────────────────────────────────────────────────────
m_nb <- glm.nb(
  species_richness ~ log_effort + season + weekend + year_c,
  data = daily
)

# ── Null model (intercept only) ───────────────────────────────────────────────
m_null <- glm(species_richness ~ 1, data = daily, family = poisson)

# ── AIC comparison ────────────────────────────────────────────────────────────
AIC_table <- tibble(
  Model  = c("Null (Poisson)", "Poisson GLM", "Negative Binomial GLM"),
  AIC    = c(AIC(m_null), AIC(m_pois), AIC(m_nb)),
  df_res = c(m_null$df.residual, m_pois$df.residual, m_nb$df.residual)
) |> arrange(AIC)

kbl(AIC_table, digits = 1,
    caption = "Table 2. AIC comparison of fixed-effects candidate models.") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r overdispersion-check}
# ── Overdispersion check ──────────────────────────────────────────────────────
pois_ratio <- m_pois$deviance / m_pois$df.residual
nb_theta   <- m_nb$theta

cat("Poisson residual deviance / df =", round(pois_ratio, 3),
    " (>1 = overdispersed; <1 = underdispersed)\n")
cat("NB theta =", round(nb_theta, 1),
    " (very large theta ≈ Poisson; ~1 = high overdispersion)\n\n")

if (pois_ratio < 1.2 & nb_theta > 100) {
  cat("➜ No meaningful overdispersion detected.\n")
  cat("  Poisson GLM is the preferred fixed-effects model.\n")
} else {
  cat("➜ Overdispersion present. Negative Binomial preferred.\n")
}

m_fixed  <- m_pois
mod_label <- "Poisson GLM"
cat("\nFinal fixed-effects model:", mod_label, "\n")
```

## Extending to a Mixed-Effects Model (GLMM)

The fixed-effects Poisson GLM treats all observations as independent. However, individual observers may have systematically different recording tendencies — some may specialise in native species, others record whatever crosses their path. Ignoring this clustering can underestimate uncertainty in fixed-effect estimates.

We fit a Poisson GLMM with the same fixed-effects structure but with **observer identity as a random intercept**, using `lme4::glmer`. This allows each observer's baseline recording rate to vary, while still estimating the population-level effects of effort, season, and time.

```{r fit-glmm}
# ── Join observer identity to daily data ─────────────────────────────────────
# For the GLMM we need at least one observer identifier per day.
# We use the modal (most common) observer on each day as a proxy.
daily_observer <- birds |>
  group_by(date) |>
  count(user_login, sort = TRUE) |>
  slice_head(n = 1) |>
  ungroup() |>
  select(date, primary_observer = user_login)

daily_glmm <- daily |>
  left_join(daily_observer, by = "date")

# ── Fit GLMM: observer as random intercept ────────────────────────────────────
# Only observers with ≥2 observation days contribute meaningful random-effect
# information; others are pooled via shrinkage. This is expected with
# opportunistic data and does not indicate model misspecification.
m_glmm <- glmer(
  species_richness ~ log_effort + season + weekend + year_c +
    (1 | primary_observer),
  data   = daily_glmm,
  family = poisson(link = "log"),
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

glmm_singular  <- isSingular(m_glmm)
glmm_re_var    <- round(as.data.frame(VarCorr(m_glmm))$vcov, 4)

cat("GLMM singular fit:", glmm_singular, "\n")
cat("Random effect variance (observer):", glmm_re_var, "\n\n")

if (glmm_singular) {
  cat("Note: Singular fit detected. Random-effect variance ≈ 0,\n")
  cat("indicating observer-level clustering is negligible once log-effort\n")
  cat("is included as a predictor. Fixed-effect estimates are equivalent\n")
  cat("to the Poisson GLM — this is an interpretable, not problematic, result.\n")
}
```

## Model Summary: Fixed Effects

```{r model-summary-fixed}
tidy_fixed <- tidy(m_fixed, exponentiate = TRUE, conf.int = TRUE)

tidy_fixed |>
  mutate(
    term = recode(term,
      "(Intercept)"    = "Intercept",
      "log_effort"     = "Log effort (observers)",
      "seasonSummer"   = "Season: Summer",
      "seasonAutumn"   = "Season: Autumn",
      "seasonWinter"   = "Season: Winter",
      "weekendWeekend" = "Weekend vs Weekday",
      "year_c"         = "Year (centred)"
    )
  ) |>
  rename(
    Term            = term,
    IRR             = estimate,
    SE              = std.error,
    z               = statistic,
    `p-value`       = p.value,
    `95% CI (low)`  = conf.low,
    `95% CI (high)` = conf.high
  ) |>
  mutate(across(where(is.numeric), ~round(.x, 3))) |>
  kbl(caption = paste0("Table 3. ", mod_label,
      " coefficients (exponentiated as Incidence Rate Ratios).
      Reference: Spring, Weekday.")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) |>
  footnote(general = "IRR = Incidence Rate Ratio. Values > 1 indicate higher expected richness relative to reference.")
```

## Model Summary: Mixed Effects (GLMM)

```{r model-summary-glmm}
tidy_glmm <- tidy(m_glmm, exponentiate = TRUE, conf.int = TRUE, effects = "fixed")

tidy_glmm |>
  filter(term != "(Intercept)") |>
  select(-effect) |>
  mutate(
    term = recode(term,
      "log_effort"     = "Log effort (observers)",
      "seasonSummer"   = "Season: Summer",
      "seasonAutumn"   = "Season: Autumn",
      "seasonWinter"   = "Season: Winter",
      "weekendWeekend" = "Weekend vs Weekday",
      "year_c"         = "Year (centred)"
    )
  ) |>
  rename(
    Term            = term,
    IRR             = estimate,
    SE              = std.error,
    z               = statistic,
    `p-value`       = p.value,
    `95% CI (low)`  = conf.low,
    `95% CI (high)` = conf.high
  ) |>
  mutate(across(where(is.numeric), ~round(.x, 3))) |>
  kbl(caption = "Table 4. Poisson GLMM fixed-effect coefficients (observer as random intercept).
      Reference: Spring, Weekday.") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) |>
  footnote(general = "Random-effect variance reported separately. IRR interpretation as above.")
```

## Coefficient Comparison: GLM vs GLMM

```{r coef-comparison, fig.width=9, fig.height=5, fig.cap="Figure 6. Incidence Rate Ratios from the Poisson GLM (circles) and Poisson GLMM with observer random effect (triangles). Consistent estimates across both models indicate that fixed-effect conclusions are robust to observer-level clustering. Red = statistically significant in GLM (p < 0.05)."}
# ── Combine GLM and GLMM estimates for comparison plot ────────────────────────
glm_coefs <- tidy_fixed |>
  filter(term != "(Intercept)") |>
  mutate(
    model = "Poisson GLM",
    term  = recode(term,
      "log_effort"     = "Log effort\n(observers)",
      "seasonSummer"   = "Summer\nvs Spring",
      "seasonAutumn"   = "Autumn\nvs Spring",
      "seasonWinter"   = "Winter\nvs Spring",
      "weekendWeekend" = "Weekend\nvs Weekday",
      "year_c"         = "Year trend"
    ),
    significant = p.value < 0.05
  )

glmm_coefs <- tidy_glmm |>
  filter(term != "(Intercept)") |>
  mutate(
    model = "Poisson GLMM\n(observer random effect)",
    term  = recode(term,
      "log_effort"     = "Log effort\n(observers)",
      "seasonSummer"   = "Summer\nvs Spring",
      "seasonAutumn"   = "Autumn\nvs Spring",
      "seasonWinter"   = "Winter\nvs Spring",
      "weekendWeekend" = "Weekend\nvs Weekday",
      "year_c"         = "Year trend"
    ),
    significant = p.value < 0.05
  )

all_coefs <- bind_rows(glm_coefs, glmm_coefs)

ggplot(all_coefs,
       aes(x = estimate,
           y = fct_reorder(term, estimate),
           colour = significant,
           shape  = model)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.2, linewidth = 0.7,
                 position = position_dodge(width = 0.5)) +
  geom_point(size = 3.5,
             position = position_dodge(width = 0.5)) +
  scale_colour_manual(values = c("FALSE" = "grey60", "TRUE" = "#e63946"),
                      labels = c("p ≥ 0.05", "p < 0.05 (GLM)"),
                      name   = "") +
  scale_shape_manual(values = c("Poisson GLM" = 16,
                                 "Poisson GLMM\n(observer random effect)" = 17),
                     name = "Model") +
  labs(
    title    = "Fixed-effect estimates: Poisson GLM vs Poisson GLMM",
    subtitle = "Observer identity as random intercept in GLMM",
    x        = "Incidence Rate Ratio (IRR)",
    y        = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "right")
```

## Model Diagnostics

```{r dharma-diagnostics, fig.width=10, fig.height=5, fig.cap="Figure 7. DHARMa residual diagnostics for the Poisson GLM. (A) QQ-plot of scaled residuals; (B) residuals vs fitted values. Approximate uniformity in (A) and no systematic trend in (B) indicate adequate model fit."}
set.seed(738)
sim_res <- simulateResiduals(fittedModel = m_fixed, n = 500, plot = FALSE)

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
plotQQunif(sim_res,    main = "A. QQ-plot (scaled residuals)")
plotResiduals(sim_res, main = "B. Residuals vs fitted")
par(mfrow = c(1, 1))
```

```{r dharma-tests}
cat("── Dispersion test ──\n")
testDispersion(sim_res)
cat("\n── Zero-inflation test ──\n")
testZeroInflation(sim_res)
```

> **Note on zero-inflation:** The zero-inflation test is significant because our `daily` table only contains days *with at least one observation*. Days with no observer activity are structurally absent from iNaturalist data — this is a feature of opportunistic sampling, not a model misspecification.

## Marginal Effect of Observer Effort

```{r marginal-effects, fig.cap="Figure 8. Predicted daily species richness as a function of observer effort, derived from the Poisson GLM. Year held at mean; Weekday. Ribbons = 95% CI. The near-doubling of predicted richness with each unit increase in log-observer count illustrates the dominance of effort over seasonal or temporal factors."}
pred_grid <- expand_grid(
  n_observers = seq(1, max(daily$n_observers), length.out = 80),
  season      = levels(daily$season),
  weekend     = "Weekday",
  year_c      = mean(daily$year_c)
) |>
  mutate(log_effort = log1p(n_observers))

preds <- predict(m_fixed, newdata = pred_grid, type = "link", se.fit = TRUE)

pred_grid <- pred_grid |>
  mutate(
    predicted = exp(preds$fit),
    lower     = exp(preds$fit - 1.96 * preds$se.fit),
    upper     = exp(preds$fit + 1.96 * preds$se.fit)
  )

ggplot(pred_grid, aes(x = n_observers, y = predicted,
                       colour = season, fill = season)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 1) +
  scale_colour_viridis_d(option = "D", direction = -1) +
  scale_fill_viridis_d(option = "D", direction = -1) +
  labs(
    title    = "Predicted species richness as a function of observer effort",
    subtitle = "Weekday, year held at mean; shaded ribbons = 95% CI",
    x        = "Number of observers",
    y        = "Predicted species richness",
    colour   = "Season", fill = "Season"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "bottom")
```

---

# Discussion

```{r discussion-values, include=FALSE}
irr_effort  <- round(tidy_fixed$estimate[tidy_fixed$term == "log_effort"], 2)
p_effort    <- tidy_fixed$p.value[tidy_fixed$term == "log_effort"]
irr_summer  <- round(tidy_fixed$estimate[tidy_fixed$term == "seasonSummer"], 2)
p_summer    <- round(tidy_fixed$p.value[tidy_fixed$term == "seasonSummer"], 3)
irr_weekend <- round(tidy_fixed$estimate[tidy_fixed$term == "weekendWeekend"], 2)
p_weekend   <- round(tidy_fixed$p.value[tidy_fixed$term == "weekendWeekend"], 3)
irr_year    <- round(tidy_fixed$estimate[tidy_fixed$term == "year_c"], 2)
p_year      <- round(tidy_fixed$p.value[tidy_fixed$term == "year_c"], 3)
n_species   <- n_distinct(birds$scientific_name)
n_obs_total <- nrow(birds)
n_days      <- nrow(daily)
```

**Can we trust the campus bird ranking?** The sensitivity analysis provides a reassuring answer. Removing the top five observers — who together account for `r top5_pct`% of all records — left the top five bird groups in the same order. This suggests the main pattern is not an artefact of a handful of enthusiastic contributors. The restaurant analogy holds: even without the power reviewers, the most popular dishes are still the same dishes. However, lower-ranked groups did shift position after removal, indicating that rare or infrequently recorded species are more sensitive to observer identity. Conclusions about common species are robust; conclusions about uncommon ones should be treated with more caution.

**What drives daily species richness?** Observer effort was by far the strongest predictor in both the fixed-effects Poisson GLM and the mixed-effects GLMM (IRR = `r irr_effort`, p < 0.001 in both models). Each unit increase in log-transformed observer count was associated with a near-doubling of expected species richness. The consistency of this estimate across the GLM and GLMM — which accounts for observer-level clustering — indicates that this is a genuine population-level signal, not an artefact of a few highly productive observers inflating the count.

Season showed no statistically significant effect (Summer: IRR = `r irr_summer`, p = `r p_summer`), nor did weekend versus weekday (IRR = `r irr_weekend`, p = `r p_weekend`) or the year trend (IRR = `r irr_year`, p = `r p_year`). The March peak visible in the monthly plot is consistent with semester timing and increased student activity rather than a genuine biological signal. When observer numbers are statistically held constant, March loses its exceptional status.

**What the GLMM adds.** The GLMM returned a singular fit, with the random-effect variance for observer identity estimated at zero. This means that once observer *count* (log-effort) is included as a predictor, knowing *which* observer was dominant on a given day explains no additional variation in species richness. Far from being a problem, this is an ecologically interpretable result: it is the number of people looking, not their individual identities, that drives how many species get recorded. The sensitivity analysis (which directly removed the most active observers) and the GLMM (which statistically adjusts for observer-level differences) arrive at the same conclusion by independent routes — strengthening confidence in the finding.

**Limitations.** The dataset is small (`r n_obs_total` observations over `r n_days` observation days), limiting statistical power. The GLMM uses the modal observer per day as a proxy for observer identity, which is a simplification when multiple observers contribute on the same day. Species identifications, while filtered to research grade, may still contain occasional errors. The campus boundary spans ecologically distinct microhabitats that are not uniformly sampled.

---

# Conclusion

Back to the restaurant. The leaderboard is trustworthy for the top dishes — take away the most active reviewers and the ranking barely budges. But for the dishes near the bottom, a single enthusiast can shift the order considerably. The same is true for the campus bird record.

More importantly: what fills a plate on a given day is not the season or the day of the week. It is how many people walked in. The birds of the UoA city campus — tūī, kererū, pīwakawaka alongside the pigeons and sparrows — are genuinely there. But how many species we record on any given day depends overwhelmingly on how many people were paying attention.

iNaturalist does not give a census. It gives a record of encounters between birds and people. That distinction matters — not as a reason to distrust the data, but as the key to reading it correctly.

The code and data supporting this analysis are available at [https://github.com/mengyuanzheng9-star/uoa-campus-biota](https://github.com/mengyuanzheng9-star/uoa-campus-biota).

---

# Ethical Considerations

## Data and Privacy

All observations are publicly available through iNaturalist under Creative Commons licences. Observer usernames were used only in aggregate (counts per day) for the fixed-effects models. For the GLMM, observer identities were used as anonymous random-effect levels to account for clustering; no attempt was made to identify, profile, or attribute behaviour to individual observers. The sensitivity analysis involved removing records by observer group, not singling out individuals.

## Interpretation of Citizen Science Data

Opportunistic iNaturalist records reflect detectability-weighted presence, not abundance or occupancy. All results are framed in terms of "recorded richness" and "observations" rather than "bird abundance" or "population size." The sensitivity analysis and modelling are designed to make the limitations of opportunistic data explicit, not to dismiss the data's value.

## AI Use Statement

Claude (Anthropic, claude-sonnet-4-6) was used to assist in generating the initial scaffold of this R Markdown document, including code structure for the GLM, GLMM, sensitivity analysis, and figure formatting. All analytical decisions — model structure, variable selection, the sensitivity analysis design, and interpretation of results — were made independently by the author. All code was reviewed, tested, and debugged by the author. The interpretation of results, Discussion, and Conclusion were written independently.

---

# References

Isaac, N.J.B., van Strien, A.J., August, T.A., de Zeeuw, M.P. & Roy, D.B. (2014). Statistics for citizen science: extracting signals of change from noisy ecological data. *Methods in Ecology and Evolution*, 5, 1052–1060.

iNaturalist (2026). Biota of University of Auckland City Campus. Available at: https://www.inaturalist.org/projects/biota-of-university-of-auckland-city-campus. Accessed 27 April 2026.

Venables, W.N. & Ripley, B.D. (2002). *Modern Applied Statistics with S*, 4th edition. Springer, New York.

Bates, D., Mächler, M., Bolker, B. & Walker, S. (2015). Fitting Linear Mixed-Effects Models Using lme4. *Journal of Statistical Software*, 67(1), 1–48.

Hartig, F. (2022). DHARMa: Residual Diagnostics for Hierarchical (Multi-Level / Mixed) Regression Models. R package version 0.4.6.

---

# Session Information

```{r session-info}
sessionInfo()
```
